--- 
title: "rOpenSci Statistical Software Peer Review"
author: "Mark Padgham and Noam Ross"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [statsoft.bib]
biblio-style: apalike
link-citations: yes
github-repo: ropenscilabs/statistical-software-review-book
url: 'https\://ropensci-blog-guidance.netlify.com/'
apple-touch-icon: "favicon/apple-touch-icon.png"
apple-touch-icon-size: 120
favicon: "favicon/favicon.ico"
---


```{r getbib, include = FALSE, eval = TRUE}
# RefManageR is no longer sufficiently reliable, so bibliography is manually
# downloaded for the moment
if (!file.exists("statsoft.bib")) {
  refs <- RefManageR::ReadZotero(
    group = "2416765",
    .params = list(limit = 100)
  )
  RefManageR::WriteBib(refs, "statsoft.bib")
}
```
```{r updates, echo = FALSE}
# https://github.com/rstudio/bookdown/issues/440
# https://bookdown.org/yihui/bookdown/html-widgets.html
install.packages("webshot")
webshot::install_phantomjs()
# install.packages("tinytex")
# tinytex::tlmgr_update()
```


# Welcome {#welcome}

Welcome to [rOpenSci](https://ropensci.org)'s project
developing [software peer review](https://ropensci.org/software-review/) system
for statistical software in R packages and  beyond.  We invite you to join us
in developing standards for processes of assessment and peer review of
statistical software.

rOpenSci currently runs software peer review for R packages focused on data
life cycle management. To apply these processes to software implementing
statistical methods and algorithms, we need new ways of evaluating and testing
software, and managing the review process. This book serves as a home for
collecting research and developing those standards. As the book develops, it
will become an extension of [rOpenSci Packages: Development, Maintenance, and
Peer Review](https://devguide.ropensci.org/), documenting not only our
guidelines for statistical software but also the process of expanding the scope
of review so it may be reproduced in other domains.

You are invited to contribute to this project by filing suggestions as
[issues in the book's GitHub repository](https://github.com/ropenscilabs/statistical-software-review-book/issues).
We also have a dedicated [area for discussion on the rOpenSci forum](https://discuss.ropensci.org/c/statistical-software-peer-review/28).
We have noted in the chapters we're actively seeking more input and feedback by
highlighting them their titles in <span style="color:red;">red and adding
`[SEEKING FEEDBACK]`</span> to them.  Feedback on other sections is also
welcome, although these may be at various stages of development.

For this project we are lucky to have support of a great steering committee.  The
committee's role is to help us with outreach to statistical communities, vetting
community input and prioritizing our work. They are:

-   [Ben Bolker](https://ms.mcmaster.ca/~bolker/) ([\@bolkerb](https://twitter.com/bolkerb)) McMaster University
-   [Rebecca Killick](http://www.lancs.ac.uk/~killick/), Lancaster University
-   [Stephanie Hicks](https://www.stephaniehicks.com/) ([\@stephaniehicks](https://twitter.com/stephaniehicks)), Johns Hopkins University
-   [Max Kuhn](http://appliedpredictivemodeling.com/) ([\@topepos](https://twitter.com/topepos)), RStudio
-   [Paula Moraga](http://www.paulamoraga.com/), King Abdullah University of Science and Technology, Saudi Arabia
-   [Leonardo Collado-Torres](http://lcolladotor.github.io/), Lieber Institute for Brain Development, USA

This work is [
supported by the Sloan Foundation](https://ropensci.org/blog/2019/07/15/expanding-software-review/)
and is organized under an
[R Consortium Working Group](https://www.r-consortium.org/projects/isc-working-groups).

## Contributors


<!-- ALL-CONTRIBUTORS-LIST:START - Do not remove or modify this section -->
<!-- prettier-ignore-start -->
<!-- markdownlint-disable -->

This project uses the [`allcontributor` package](https://github.com/mpadge/allcontributor) following the [all-contributors](https://allcontributors.org) specification. Contributions of any kind are welcome!

## Code

<table>

<tr>
<td align="center">
<a href="https://github.com/mpadge">
<img src="https://avatars1.githubusercontent.com/u/6697851?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/commits?author=mpadge">mpadge</a>
</td>
<td align="center">
<a href="https://github.com/noamross">
<img src="https://avatars1.githubusercontent.com/u/571752?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/commits?author=noamross">noamross</a>
</td>
<td align="center">
<a href="https://github.com/maelle">
<img src="https://avatars0.githubusercontent.com/u/8360597?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/commits?author=maelle">maelle</a>
</td>
<td align="center">
<a href="https://github.com/richfitz">
<img src="https://avatars1.githubusercontent.com/u/1558093?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/commits?author=richfitz">richfitz</a>
</td>
<td align="center">
<a href="https://github.com/elong0527">
<img src="https://avatars0.githubusercontent.com/u/465246?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/commits?author=elong0527">elong0527</a>
</td>
<td align="center">
<a href="https://github.com/Pakillo">
<img src="https://avatars0.githubusercontent.com/u/1597843?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/commits?author=Pakillo">Pakillo</a>
</td>
</tr>

</table>


## Issue Authors

<table>

<tr>
<td align="center">
<a href="https://github.com/elimillera">
<img src="https://avatars1.githubusercontent.com/u/24767886?u=16d22ed97b442cfa1847d0f110b7dbf5c190c5a5&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/issues?q=is%3Aissue+author%3Aelimillera">elimillera</a>
</td>
<td align="center">
<a href="https://github.com/IndrajeetPatil">
<img src="https://avatars0.githubusercontent.com/u/11330453?u=6eb9f03fe4cc5a2a67633252acda215b113234be&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/statistical-software-review-book/issues?q=is%3Aissue+author%3AIndrajeetPatil">IndrajeetPatil</a>
</td>
</tr>

</table>

<!-- markdownlint-enable -->
<!-- prettier-ignore-end -->
<!-- ALL-CONTRIBUTORS-LIST:END -->
