--- 
title: "rOpenSci Statistical Software Peer Review"
author: "Mark Padgham and Noam Ross"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [statsoft.bib]
biblio-style: apalike
link-citations: yes
github-repo: ropensci/statistical-software-review-book
url: 'https\://ropensci-blog-guidance.netlify.com/'
apple-touch-icon: "favicon/apple-touch-icon.png"
apple-touch-icon-size: 120
favicon: "favicon/favicon.ico"
---


```{r getbib, include = FALSE, eval = TRUE}
# RefManageR is no longer sufficiently reliable, so bibliography is manually
# downloaded for the moment
if (!file.exists("statsoft.bib")) {
  refs <- RefManageR::ReadZotero(
    group = "2416765",
    .params = list(limit = 100)
  )
  RefManageR::WriteBib(refs, "statsoft.bib")
}
```
```{r updates, echo = FALSE}
# https://github.com/rstudio/bookdown/issues/440
# https://bookdown.org/yihui/bookdown/html-widgets.html
#install.packages("webshot")
webshot::install_phantomjs()
# install.packages("tinytex")
# tinytex::tlmgr_update()
```


# Welcome {#welcome}

Welcome to [rOpenSci](https://ropensci.org)'s system for peer-review of
statistical software in R packages and beyond. This system extends our existing
system for [software peer
review](https://ropensci.org/software-review/), through [expanding the
scope](https://ropensci.org/blog/2019/07/15/expanding-software-review/) to
include explicitly statistical software. As such, it is a direct extension of
[rOpenSci Packages: Development, Maintenance, and
Peer Review](https://devguide.ropensci.org/). This book provides guidelines for
authors on how to develop statistical software, and for editors and reviewers
on our processes for peer review of statistical software.

You are invited to contribute to this project by filing suggestions as [issues
in the book's GitHub
repository](https://github.com/ropensci/statistical-software-review-book/issues).
We also have a dedicated [area for discussion on the rOpenSci
forum](https://discuss.ropensci.org/c/statistical-software-peer-review/28).
Feedback on standards is particularly welcome.

For this project we are lucky to have support of a great advisory and editorial
committee.  The committee has helped us along with way with outreach to
statistical communities, vetting community input and prioritizing our work.

-   [Ben Bolker](https://ms.mcmaster.ca/~bolker/)
    ([\@bolkerb](https://twitter.com/bolkerb)) McMaster University, Canada
-   [Rebecca Killick](http://www.lancs.ac.uk/~killick/), Lancaster University, UK
-   [Stephanie Hicks](https://www.stephaniehicks.com/)
  ([\@stephaniehicks](https://twitter.com/stephaniehicks)), Johns Hopkins
  University, USA
-   [Paula Moraga](http://www.paulamoraga.com/), King Abdullah University of
  Science and Technology, Saudi Arabia
-   [Leonardo Collado-Torres](http://lcolladotor.github.io/)
  ([\@lcolladotor](https://twitter.com/lcolladotor)), Lieber Institute for
  Brain Development, USA
-   [Toby Hocking](http://tdhock.github.io/), Northern Arizona University, USA

We are also grateful to former members of our advisory board:

-   [Max Kuhn](http://appliedpredictivemodeling.com/)
    ([\@topepos](https://twitter.com/topepos)), RStudio
-   [Martin Morgan](https://www.roswellpark.org/martin-morgan)
    ([\@mt_morgan](https://twitter.com/mt_morgan)), Roswell Park Comprehensive
    Cancer Center



This work is [ supported by the Sloan
Foundation](https://ropensci.org/blog/2019/07/15/expanding-software-review/)
and is organized under an [R Consortium Working
Group](https://www.r-consortium.org/projects/isc-working-groups).

## Citation

You can cite this book by its [its Zenodo
DOI](https://doi.org/10.5281/zenodo.5556755), or by copying the following
BibTeX entry:


```{r zenodo, echo=FALSE, results = 'asis', warning=FALSE}
cat("\n```bibtex\n")
u <- "https://zenodo.org/record/5556755" # universal URL redirects to latest
s <- rvest::session(u)
u <- paste0(s$url, "/export/hx") # redirected
out <- rvest::read_html(u)
bibentry <- rvest::html_element(out, "pre")
out <- strsplit (rvest::html_text(bibentry), "\n")[[1]]
writeLines(out, "dev_citation_BibTeX.bib")
cat (rvest::html_text (bibentry))
cat("\n```\n")
```

## Contributors






<!-- ALL-CONTRIBUTORS-LIST:START - Do not remove or modify this section -->
<!-- prettier-ignore-start -->
<!-- markdownlint-disable -->

All contributions to this project are gratefully acknowledged using the [`allcontributors` package](https://github.com/ropenscilabs/allcontributors) following the [all-contributors](https://allcontributors.org) specification. Contributions of any kind are welcome!

### Content

<table>

<tr>
<td align="center">
<a href="https://github.com/mpadge">
<img src="https://avatars.githubusercontent.com/u/6697851?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=mpadge">mpadge</a>
</td>
<td align="center">
<a href="https://github.com/noamross">
<img src="https://avatars.githubusercontent.com/u/571752?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=noamross">noamross</a>
</td>
<td align="center">
<a href="https://github.com/maelle">
<img src="https://avatars.githubusercontent.com/u/8360597?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=maelle">maelle</a>
</td>
<td align="center">
<a href="https://github.com/Nowosad">
<img src="https://avatars.githubusercontent.com/u/3457131?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=Nowosad">Nowosad</a>
</td>
<td align="center">
<a href="https://github.com/richfitz">
<img src="https://avatars.githubusercontent.com/u/1558093?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=richfitz">richfitz</a>
</td>
</tr>


<tr>
<td align="center">
<a href="https://github.com/elong0527">
<img src="https://avatars.githubusercontent.com/u/465246?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=elong0527">elong0527</a>
</td>
<td align="center">
<a href="https://github.com/Pakillo">
<img src="https://avatars.githubusercontent.com/u/1597843?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=Pakillo">Pakillo</a>
</td>
<td align="center">
<a href="https://github.com/briatte">
<img src="https://avatars.githubusercontent.com/u/322533?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=briatte">briatte</a>
</td>
<td align="center">
<a href="https://github.com/lcolladotor">
<img src="https://avatars.githubusercontent.com/u/2288213?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/commits?author=lcolladotor">lcolladotor</a>
</td>
</tr>

</table>


### Issue Authors

<table>

<tr>
<td align="center">
<a href="https://github.com/elimillera">
<img src="https://avatars.githubusercontent.com/u/24767886?u=16d22ed97b442cfa1847d0f110b7dbf5c190c5a5&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/issues?q=is%3Aissue+author%3Aelimillera">elimillera</a>
</td>
<td align="center">
<a href="https://github.com/IndrajeetPatil">
<img src="https://avatars.githubusercontent.com/u/11330453?u=6eb9f03fe4cc5a2a67633252acda215b113234be&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/issues?q=is%3Aissue+author%3AIndrajeetPatil">IndrajeetPatil</a>
</td>
<td align="center">
<a href="https://github.com/jsta">
<img src="https://avatars.githubusercontent.com/u/7844578?u=a37c1879ee6da8d66d2cf6f7b79d08bcb837cbea&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci/statistical-software-review-book/issues?q=is%3Aissue+author%3Ajsta">jsta</a>
</td>
</tr>

</table>

<!-- markdownlint-enable -->
<!-- prettier-ignore-end -->
<!-- ALL-CONTRIBUTORS-LIST:END -->
