```{r include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE
)
```
```{r include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE
)
```

--- 
title: "rOpenSci Statistial Software Peer Review"
author: "Mark Padgham and Noam Ross"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [statsoft.bib]
biblio-style: apalike
link-citations: yes
github-repo: ropenscilabs/statistical-software-peer-review
url: 'https\://ropensci-blog-guidance.netlify.com/'
apple-touch-icon: "favicon/apple-touch-icon.png"
apple-touch-icon-size: 120
favicon: "favicon/favicon.ico"
---


```{r getbib, include = FALSE, eval = TRUE}
if (!file.exists ("statsoft.bib")) {
   refs <- RefManageR::ReadZotero(group = "2416765",
                                 .params = list (limit = 100))
   RefManageR::WriteBib(refs, "statsoft.bib")
}
```

# Welcome {#welcome}

Welcome to [rOpenSci](https://ropensci.org)'s project
developing [software peer review](https://ropensci.org/software-review/) system
for statistical software in R packages and  beyond.  We invite you to join us in
developing standards for peer-reviewed evaluation of statistical software.

rOpenSci runs software peer review for R packages focused on data lifecycle management.
To apply these processes to software implementing statistical methods and algorithms,
we need new ways of evaluating and testing software, and managing the review
process. This book serves as a home for collecting research and and developing
those standards. As the book develops, it will become an extension of
[rOpenSci Packages: Development, Maintenance, and Peer Review](https://devguide.ropensci.org/),
documenting not only our guidelines for statistical software but also the process
of expanding the scope of review may be reproduced in other domains.

You are invited to contribute to this project by filing suggestions as
[issues in the book's GitHub repository](https://github.com/ropenscilabs/statistical-software-peer-review/issues).
We also have a dedicated [area for discussion on the rOpenSci forum](https://discuss.ropensci.org/c/statistical-software-peer-review/28).

For this project we are lucky to have support of a great steering committee.  The
commitee's role is to help us with outreach to statistical communities, vetting
community input and prioritizing our work. They are:

-   [Ben Bolker](https://ms.mcmaster.ca/~bolker/) ([\@bolkerb](https://twitter.com/bolkerb)) McMaster University
-   [Rebecca Killick](http://www.lancs.ac.uk/~killick/), Lacncaster University
-   [Stephanie Hicks](https://www.stephaniehicks.com/) ([\@stephaniehicks](https://twitter.com/stephaniehicks)), Johns Hopkins University
-   [Max Kuhn](http://appliedpredictivemodeling.com/) ([\@topepos](https://twitter.com/topepos)), RStudio
-   [Martin Morgan](https://www.roswellpark.org/martin-morgan) ([\@mt_morgan](https://twitter.com/mt_morgan)), Roswell Park Comprehensive Cancer Center

This work is [
supported by the Sloan Foundation](https://ropensci.org/blog/2019/07/15/expanding-software-review/)
and is organized under an
[R Consortium Working Group](https://www.r-consortium.org/projects/isc-working-groups).
