#  (PART) Package Development, Submission, and Maintenance {-}


# Package Development {#pkgdev}

Package developers are expected to follow the three sub-sections described in
this chapter, starting with a determination of whether a package is likely to
be considered within the scope of our review system. We then describe
our automated assessment tools, which packages are expected to pass, followed
by a description of our system for which developers are expected to self-assess
their software against all applicable standards. As described at the outset,
this book is also the definitive reference for all standards, as provided in
the [dedicated chapter](#standards).

## Scope

The first important task prior to submitting a package is to estimate whether
a package is likely to be considered within our scope for statistical software.
As described in the [Overview](#overview-scope), packages are generally
considered in scope if they fit one or more of the [categories listed
there](#overview-categories), and in the corresponding standards.

Developers are encouraged to contact us at any point prior to, or during,
development, to ask about whether a package might be in scope, or which
categories it might fit within. Categorisation of packages may not always be
straightforward, and we particularly encourage developers who are unsure about
whether a package belongs in a particular category or not to contact us for
discussion. An initial judgement of whether or not a package belongs in
a particular category may be gained by examining the respective standards. Any
package for which a large number of standards from a particular category may be
considered applicable (regardless of whether or not they would actually be
checked) is likely to fit within that category. Once you have determined that
your package is likely to fit into one or more of our in-scope categories,
you'll need to apply our two primary development tools described in the
following two sub-sections.

## The [`autotest` package](https://github.com/ropenscilabs/autotest) {#pkgdev-autotest}

We have developed an automated assessment tool,
[`autotest`](https://github.com/ropenscilabs/autotest), which all packages are
expected to pass in order to be accepted for submission. The package implements
a form of "mutation testing," by examining the types of all input parameters,
implementing type-specific mutations, and examining the response of each
function in a package to all such mutations. This kind of mutation testing is
a very effective way to uncover any unexpected behaviour which developers
themselves might not necessarily pre-empt.

Full documentation is provided on the [package
website](https://ropenscilabs.github.io/autotest/), and we particularly
encourage any developers intending to develop packages for submission to our
peer review system to step through the [main `autotest`
vignette](https://ropenscilabs.github.io/autotest/articles/autotest.html), and
to apply [`autotest`](https://ropenscilabs.github.io/autotest/) continuously
throughout package development, to ensure that `autotest_package()` returns
clean (`NULL`) results when the package is first submitted.


## Assessment Against Standards

Once a package has been sufficiently developed to begin alignment with our
standards, and once all issues revealed by
[`autotest`](https://ropenscilabs.github.io/autotest/) have been addressed,
developers will need to use our second tool, the [`ssr` (**s**oftware
**r**eveiw **r**oclets) package](https://ropenscilabs.github.io/srr/) to insert
both general and category-specific standards into their code, and to begin the
process of documenting within the code itself how and where the code adheres to
the individual standards.

The procedures are again described in detail on the [main `srr`
website](https://ropenscilabs.github.io/srr/), and in particular in the
[package vignette](https://ropenscilabs.github.io/srr/articles/srr-stats.html).
Developers are first encouraged to obtain a local copy of the [source code for
that
vignette](https://github.com/ropenscilabs/srr/blob/main/vignettes/srr-stats.Rmd),
and to step through each line in order to understand how the procedure works.
Having done that, you may then insert standards into your own package by
running from within the local directory of your package,

```{r ssr-standards, eval = FALSE, echo = TRUE}
srr_stats_roxygen (category = c ("<category-1>", "<category-2>"))
```

That will insert a new file into the `R/` directory of your package called (by
default) `srr-stats-standards.R`. All standards initially have
a [`roxygen2`](https://roxygen2.r-lib.org) tag of `@srrstatsTODO`, to indicate
that these standards are yet to be addressed. These tags are processed by the 
[`srr` roclet](https://ropenscilabs.github.io/srr/) which needs to be connected
with your package by modifying the `Roxygen` line of your `DESCRIPTION` file to
the following form:

```{r srr-DESC, eval = FALSE, echo = TRUE}
Roxygen: list (markdown = TRUE, roclets = c ("namespace", "rd", "srr::srr_stats_roclet"))
```

You do not need to add the `srr` package anywhere else in your `DESCRIPTION`
file, nor do you need to retain this line when submitting packages to CRAN (or
elsewhere). You should nevertheless retain the line at all other times, and you
can easily disable the roclet output by including `#' @srrVerbose FALSE`
somewhere within your documentation. Note that `srr` documentation lines are
used only to produce on-screen output triggered by running
[`roxygen2::roxygensise()`](https://roxygen2.r-lib.org/reference/roxygenize.html),
or the equivalent function,
[`devtools::document()`](http://devtools.r-lib.org/reference/document.html),
and do not appear in actual package documentation.

The [`srr` roclet](https://ropenscilabs.github.io/srr/) recognises and process
three tags:

1. `@srrstatsTODO` to flag standards yet to be addressed;
2. `@srrstats` to flag standards which have been addressed, and followed by
   descriptions of how your code addresses those standards; and
3. `@srrstatsNA` to flag standards which you deem not to be applicable to your
   code, followed by explanations of why you deem those standards not
   applicable.

The file generated by
[`srr_stats_roxygen()`](https://ropenscilabs.github.io/srr/reference/srr_stats_roxygen.html)
initially contains two [`roxygen2`](https://roxygen2.r-lib.org) blocks, the
first containing every standard potentially applicable to your package, tagged
with `@srrstatsTODO`, and the second with a title of `MA_standards`, to
document standards deemed not applicable.

The first task after having generated this file is to move standards to
approximate locations within your package where they are likely to be
addressed. For example, standards concerning tests should be moved somewhere
within the `tests/` directory, standards concerning documentation to the main
`README.Rmd` file, or within a vignette file. The [package
skeleton](https://ropenscilabs.github.io/srr/reference/srr_stats_pkg_skeleton.html)
includes code demonstrating how to include roclet tags within `.Rmd` files.
This will break down an initially large single list of standards into more
manageable groups of standards dispersed throughout your code. As each standard
is addressed, it should be moved to one or more locations in your code as near
as possible to relevant code, and the tag changed from `@srrstatsTODO` to
`@srrstats`, and a brief description appended to explain how that standard is
addressed.

Standards deemed not to be applicable to your package should all be group together within a single 
[`roxygen2`](https://roxygen2.r-lib.org) block with a title of `NA_standards`,
each with a tag of `@srrstatsNA`, and a brief description of why those
standards are deemed not to be applicable. Full details of the `srr` procedure
are given on the [pakcage website](https://ropenscilabs.github.io/srr)/
