# Standards: Version {{< var version >}} {#standards}


This Chapter serves as the reference for rOpenSci's standards for statistical
software. Software accepted for peer-review must fit one or more of our
categories, and thus all packages must comply with the *General Standards*
listed in the first of the following sections, along with one or more of the
category-specific sets of standards listed in the subsequent sections.

Our standards are open and intended to change and evolve in response to public
feedback. Please contribute via the [GitHub discussions pages for this
book](https://github.com/ropensci/statistical-software-review-book/discussions).
We particularly encourage anybody preparing software for submission to discuss
any aspects of our standards, including applicability, validity, phrasing,
expectations, reasons for standards, and even the addition or removal of
specific standards.


---

```{r standards-gen}
#| eval: true
#| echo: false
#| child: "standards/general.Rmd"
```

---

```{r standards-bayes}
#| eval: true
#| echo: false
#| child: "standards/bayesian.Rmd"
```

---

```{r standards-eda}
#| eval: true
#| echo: false
#| child: "standards/eda.Rmd"
```

---

```{r standards-ml}
#| eval: true
#| echo: false
#| child: "standards/ml.Rmd"
```

---

```{r standards-regression}
#| eval: true
#| echo: false
#| child: "standards/regression.Rmd"
```

---

```{r standards-spatial}
#| eval: true
#| echo: false
#| child: "standards/spatial.Rmd"
```

---


```{r standards-ts}
#| eval: true
#| echo: false
#| child: "standards/time-series.Rmd"
```

---

```{r standards-unsupervised}
#| eval: true
#| echo: false
#| child: "standards/unsupervised.Rmd"
```
