
#  (PART) Standards {-}

# Standards: Version 0.0.X {#standards}

This Chapter serves as the reference for rOpenSci's standards for statistical
software. Software accepted for peer-review must fit one or more of our
categories, and thus all packages must comply with the *General Standards*
listed in the first of the following sections, along with one or more of the
category-specific sets of standards listed in the subsequent sections.

Examples of application of these standards may be viewed as separate
[`hackmd.io` files](https://hackmd.io) by clicking on the following links:

- [Application of Bayesian and Monte Carlo Standards](https://hackmd.io/zVWTAl9ZQeCcj_bvMGcmMQ)
- [Application of Regression and Supervised Learning Standards](https://hackmd.io/VZ-wgQtZRV2pb-wFZNDM5g)
- [Application of Dimensionality Reduction, Clustering, and Unsupervised Learning Standards](https://hackmd.io/iOZD_oCpT86zoY5z4memaQ)
- [Application of Exploratory Data Analysis Standards](https://hackmd.io/K8F1RIhdQeuZFqMnzdqNVw)
- [Application of Machine Learning Software Standards](https://hackmd.io/Ix1YwD8YTWGuzdiXsVQadA)

Each of those files compares both general and category-specific standards
against selected R packages within those categories. These comparisons are
intended for illustrative purposes only, and are in no way intended to
represent evaluations of the software. They are presented in the hope of
demonstrating how the standards presented here may be applied to software, and
what the results of such application may look like.

```{r standards-gen, child = "standards/general.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-bayes, child = "standards/bayesian.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-eda, child = "standards/eda.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-ml, child = "standards/ml.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-reg, child = "standards/regression.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-sp, child = "standards/spatial.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-ts, child = "standards/time-series.Rmd", eval = TRUE, echo = TRUE}
```

```{r standards-unsup, child = "standards/unsupervised.Rmd", eval = TRUE, echo = TRUE}
```
