---
title: "Examples of Categorizing Abstracts"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: false
        number_sections: false
        theme: flatly
---

```{r options, echo = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  fig.path="figures/"
)
```

This document explores example categorizations of abstracts from software or
associated manuscripts submitted both to rOpenSci and the Journal of Open
Source Software (JOSS). Categories are considered in two distinct domains:

1. Categories of statistical software; and
2. Categories of statistical standards that may be applied

## Example Categorizations

The categorizations that follow are illustrative only, and are not intended to
be definitive statements of categorization. Rather, they are intended merely to
illustrates the kinds of categories which may be usefully distinguished and
applied.

software | forum | software categories | standards categories
--- | --- | --- | ---
[`spatialwarnings`](https://github.com/ropensci/software-review/issues/197) | rOpenSci | workflow | numerical, algorithm, method validity, summary stats, viz
[`heatwavers`](https://github.com/ropensci/software-review/issues/219) | rOpenSci | algorithm | numerical, algorithm, method validiy
[`hhi`](https://github.com/ropensci/software-review/issues/275) | rOpenSci | algorithm | numerical, algorithm, viz
[`economiccomplexity`](https://github.com/ropensci/software-review/issues/312) | rOpenSci | algorithm, wrapper | algorithm, method validigy
[`windfarmGA`](https://github.com/ropensci/software-review/issues/331) | rOpenSci | (genetic) algorithm | algorithm
[`gtsummary`](https://github.com/ropensci/software-review/issues/334) | rOpenSci | workflow, summary stats | summary stats
[`ivis`](https://joss.theoj.org/papers/10.21105/joss.01596) | JOSS | (neural network) algorithm | algorithm, data, method
[`greta`](https://joss.theoj.org/papers/10.21105/joss.01601) | JOSS | wrapper, workflow, summary stats | inference, summary stats
[`survPen`](https://joss.theoj.org/papers/10.21105/joss.01434) | JOSS | algorithm | algorithm, numerical, methods
[`smartEDA`](https://joss.theoj.org/papers/10.21105/joss.01509) | JOSS | workflow, summary stats | summary stats
[`DscoreApp`](https://joss.theoj.org/papers/10.21105/joss.01764) | JOSS | algorithm | numerical, algorith, method validity, viz
[`modelStudio`](https://joss.theoj.org/papers/10.21105/joss.01798) | JOSS | workflow, summary stats | summary stats, viz
[`tabula`](https://joss.theoj.org/papers/10.21105/joss.01821) | JOSS | workflow, summary stats | summary stats, data

That suffices to estimate what a cross-tabulation of software versus standards
categories might look like:

Software (down) / Standards (across) | viz | numerical | algorithm | method validity | data | summary stats | inference
--- | --- | --- | --- | --- | --- | --- | --- | ---
algorithm | 2 | 4 | 7 | 5 | 1 | - | -
workflow | 3 | 1 | 1 | 1 | 1 | 6 | 1
wrapper | - | - | 1 | 1 | - | 1 | 1
summary stats | 1 | - | - | - | 1 | 5 | 1

## Summary

The preceding explorations were intended to reveal what form a potential
cross-tabulation like the above might take. The primary conclusion would seem
to be that such a table can be conceived of and constructed in a roughly
diagonal form (such that the highest entries are on or near the diagonal), yet
that almost any off-diagonal cells may also contain non-zero values. There is
thus a strong relationship between software and statistical categories, yet the
two do differ in potentially important ways. Notable exceptions in this toy
example appear to arise in regard to standards for *data* and *visualization*.

That table is also useful to indicate the relative commonality of different
categories of software, noting that single software packages may encompass
multiple categories. The *algorithm* category arose 19 times; the *workflow*
category 14 times, the *summary stats* category 8 times, and the *wrapper*
category 4 times.


# All JOSS Abstracts

Attempted categorisation of *all* JOSS abstracts which might potentially come
within scope (excepting a number of packages which implement statistical
algorithms exclusively to analyse genetic or other molecular sequences, or
(phylo-)genetic trees). The following acronyms are employed:

- "EDA": Exploratory Data Analysis
- "ML": Machine Learning

(Note that numbers correspond to the full list of JOSS papers extracted in
[`get-joss-papers.Rmd`](https://github.com/ropenscilabs/statistical-software/blob/master/abstracts/get-joss-papers.Rmd#L135-L140),
but are markdown-converted here to sequential numbers. For original
enumeration, see the `joss-papersmd-all.html` document produced by running the
entire [`get-joss-papers.Rmd`
script](https://github.com/ropenscilabs/statistical-software/blob/master/abstracts/get-joss-papers.Rmd).)

4. [`simode`](https://joss.theoj.org/papers/10.21105/joss.01850): inference, dynamical systems, likelihood, differential equations; input: equations; output: parameter estimates
5. [`mlr3`](https://joss.theoj.org/papers/10.21105/joss.01903): ML, workflow; input: parameters; output: trained ML model
6. [`perccalc`](https://joss.theoj.org/papers/10.21105/joss.01796): categorical variables, estimates; input: data.frame; output: numeric vectors, parameters
7. [`xrnet`](https://joss.theoj.org/papers/10.21105/joss.01761): regression, high-dimensional data; input: data.frame, parameters; output: model estimates
8. [`univariateML`](https://joss.theoj.org/papers/10.21105/joss.01863): estimates, maximum liklihood, probability density, probability distributions; input: numeric vector; output: model estimates
12. [`tabula`](https://joss.theoj.org/papers/10.21105/joss.01821): applied statistics, chronology, discrete data, index; input: matrix; output: summary statistics, visualization
15. [`memochange`](https://joss.theoj.org/papers/10.21105/joss.01820): time series, non-stationarity; input: numeric vector, time series; output: summary statistics, model estimates
16. [`hopit`](https://joss.theoj.org/papers/10.21105/joss.01508): discrete data, ordered response data, reporting, summary statistics; input: data.frame, categorical data; output: model estimates, summary statistics
18. [`modelStudio`](https://joss.theoj.org/papers/10.21105/joss.01798): ML, workflow, EDA; input: data.frame, ML data; output: model prediction, visualization
21. [`DscoreApp`](https://joss.theoj.org/papers/10.21105/joss.01764): categorical variables, estimates, error estimates; input: data.frame, application-specific data format; output: summary statistics, visualization
23. [`thurstonianIRT`](https://joss.theoj.org/papers/10.21105/joss.01662): latent variables, categorical variables, pairwise comparisons, model probability; input: data.frame, categorical data; output: model estimates, summary statistics
24. [`molic`](https://joss.theoj.org/papers/10.21105/joss.01665): outlier detection, high-dimensional data, networks; input: application-specific data format; output: summary statistics, model estimates
27. [`kdensity`](https://joss.theoj.org/papers/10.21105/joss.01566): non-parametric, density estimator, probability density, probability distributions, kernel density, maximum likelihood; input: data.frame, numeric vector; output: S3 function
30. [`ccostr`](https://joss.theoj.org/papers/10.21105/joss.01593): non-stationarity, estimates, survival; input: numeric vector; output: summary statistics
31. [`SmartEDA`](https://joss.theoj.org/papers/10.21105/joss.01509): EDA, workflow, variable selection, descriptive statistics, information statistics, summary statistics; input: numeric vectors; output: summary statistics
32. [`modeLLtest`](https://joss.theoj.org/papers/10.21105/joss.01542): model selection, cross-validation, likelihood; input: data.frame, formula; output: summary statistics
33. [`survPen`](https://joss.theoj.org/papers/10.21105/joss.01434): survival, regression, parameter estimation, multidimensional model, splines; input: numeric vector; output: model estimates
37. [`bayestestR`](https://joss.theoj.org/papers/10.21105/joss.01541): Bayesian, null-hypothesis testing; input: numeric vectors, formula; output: summary statistics
38. [`greta`](https://joss.theoj.org/papers/10.21105/joss.01601): Bayesian, Monte Carlo, likelihood; input: numeric vectors, formula; output: summary statistics, model prediction
39. [`ivis`](https://joss.theoj.org/papers/10.21105/joss.01596): dimensionality reduction, neural network, ML; input: data.frame; output: model estimates
41. [`lfda`](https://joss.theoj.org/papers/10.21105/joss.01572): dimensionality reduction; input: data.frame; output: model estimates
42. [`tcherry`](https://joss.theoj.org/papers/10.21105/joss.01480): trees, networks, maximum likelihood, categorical variables; input: data.frame; output: network, adjacency matrix
43. [`fmcmc`](https://joss.theoj.org/papers/10.21105/joss.01427): Monte Carlo; input: numeric vector, data.frame; output: model estimates, summary statistics
45. [`R-fitODBOD`](https://joss.theoj.org/papers/10.21105/joss.01505): binomial distribution, over-dispersion; input: data.frame; output: summary statistics
46. [`qtl2pleio`](https://joss.theoj.org/papers/10.21105/joss.01435): gene loci, random effects, generalized least squares; input: genetic data; output: summary statistics
47. [`modelDown`](https://joss.theoj.org/papers/10.21105/joss.01444): ML, summary statistics, graphical output; input: ML data; output: summary web page
48. [`insight`](https://joss.theoj.org/papers/10.21105/joss.01412): statistical models, summary statistics, algorithm choice; input: data.frame, numeric vector; output: summary statistics
52. [`ReinforcementLearning`](https://joss.theoj.org/papers/10.21105/joss.01087): reinforcement, ML, sampling, model strategy, dimensionality reduction; input: data.frame, numeric vectors; output: model estimates, summary statistics
55. [`tsfeaturex`](https://joss.theoj.org/papers/10.21105/joss.01279): time series, dimensionality reduction, feature selection; input: time series; output: summary statistics, model estimates
58. [`areal`](https://joss.theoj.org/papers/10.21105/joss.01221): interpolation, areal weights; input: spatial data; output: model estimates
64. [`riskclustr`](https://joss.theoj.org/papers/10.21105/joss.01269): risk, clustering, dimensionality reduction; input: data.frame, numeric vectors; output: summary statistics, model estimates
65. [`ChiRP`](https://joss.theoj.org/papers/10.21105/joss.01287): Monte Carlo, regression, clustering; input: data.frame, numeric vectors; output: model estimates
66. [`klrfome`](https://joss.theoj.org/papers/10.21105/joss.00722): kernel density, kernel logistic regression, spatial, areal statistics, similarity statistics, focal windows; input: data.frame; output: summary statistics
75. [`BoltzMM`](https://joss.theoj.org/papers/10.21105/joss.01193): ML, probability density, probability distributions, matrix algebra; input: binary vectors; output: summary statistics, probability
76. [`polyCub`](https://joss.theoj.org/papers/10.21105/joss.01056): integration, cubature; input: numeric vector; output: summary statistics
80. [`mbir`](https://joss.theoj.org/papers/10.21105/joss.00746): inference, effect sizes, test selection; input: numeric vectors; output: statistical parameters, summary statistics
82. [`arviz`](https://joss.theoj.org/papers/10.21105/joss.01143): Bayesian, model selection, EDA; input: ML data, data.frame; output: summary statistics
85. [`overlapping`](https://joss.theoj.org/papers/10.21105/joss.01023): kernel density, probability distributions; input: numeric vectors; output: summary statistics
86. [`iRF`](https://joss.theoj.org/papers/10.21105/joss.01077): random forests, dimensionality reduction; input: ML data; output: model estimates
89. [`qsort`](https://joss.theoj.org/papers/10.21105/joss.00911): categorical variables; input: application-specific data format; output: model estimates
90. [`ssdtools`](https://joss.theoj.org/papers/10.21105/joss.01082): probability distributions, information statistics, confidence intervals, maximum likelihood; input: data.frame; output: model estimates
93. [`gravity`](https://joss.theoj.org/papers/10.21105/joss.01038): probability distributions, regression; input: data.frame; output: model estimates, summary statistics
96. [`survxai`](https://joss.theoj.org/papers/10.21105/joss.00961): ML, regression, model selection, survival, visualization; input: ML data; output: summary statistics, reports
99. [`rr2`](https://joss.theoj.org/papers/10.21105/joss.01028): correlation, variance; input: numeric vectors; output: summary statistics
100. [`dml`](https://joss.theoj.org/papers/10.21105/joss.01036): ML, distance metrics; input: data.frame; output: summary statistics
102. [`data.adapt.multi.test`](https://joss.theoj.org/papers/10.21105/joss.00161): inference, effect sizes, dimensionality reduction; input: data.frame, ML data; output: summary statistics
103. [`compboost`](https://joss.theoj.org/papers/10.21105/joss.00967): dimensionality reduction, feature selection, regression, splines, visualization, survival, functional data analysis; input: data.frame; output: model prediction, summary statistics
104. [`multistateutils`](https://joss.theoj.org/papers/10.21105/joss.00893): discrete data, multi-state model, survival; input: numeric & categorical variables; output: model estimates, model predictions
110. [`ungroup`](https://joss.theoj.org/papers/10.21105/joss.00937): histograms, grouped data, latent variables, regression; input: binned data; output: model estimates
113. [`blendR`](https://joss.theoj.org/papers/10.21105/joss.00886): sampling, probability density, probability distributions, estimates, bias; input: numeric vectors; output: summary statistics
116. [`TDAstats`](https://joss.theoj.org/papers/10.21105/joss.00860): dimensionality reduction, feature selection, workflow; input: data.frame, numeric variables; output: model estimates, summary statistics
119. [`hhi`](https://joss.theoj.org/papers/10.21105/joss.00828): index, visualization, overlap; input: data.frame, numeric variables; output: summary statistics, visualization
120. [`logKDE`](https://joss.theoj.org/papers/10.21105/joss.00870): EDA, density estimator, kernel density; input: data.frame, numeric variables; output: model estimates
124. [`ggeffects`](https://joss.theoj.org/papers/10.21105/joss.00772): regression, interaction terms, marginal effects; input: data.frame, numeric variables; output: summary statistics, model predictions
125. [`iml`](https://joss.theoj.org/papers/10.21105/joss.00786): ML, feature selection; input: ML data; output: model estimates
127. [`rsimsum`](https://joss.theoj.org/papers/10.21105/joss.00739): Monte Carlo, estimates, simulation; input: data.frame, ML data; output: summary statistics, visualization
129. [`philentropy`](https://joss.theoj.org/papers/10.21105/joss.00765): similarity statistics, distance metrics; input: numeric variables, data.frame; output: summary statistics
130. [`disclapmix`](https://joss.theoj.org/papers/10.21105/joss.00748): probability distributions; input: application-specific data; output: probability estimates, model estimates, summary statistics
131. [`qicharts2`](https://joss.theoj.org/papers/10.21105/joss.00699): networks, stationarity; input: data.frame; output: summary statistics, visualization
132. [`scanstatistics`](https://joss.theoj.org/papers/10.21105/joss.00515): spatial, clustering; input: data.frame; output: summary statistics
136. [`humanleague`](https://joss.theoj.org/papers/10.21105/joss.00629): synthetic data, probability distributions, sampling; input: parameters; output: populations
140. [`autoplotly`](https://joss.theoj.org/papers/10.21105/joss.00657): workflow, data transformation, visualization; input: data.frame, numeric variables; output: visualization
141. [`MCMCvis`](https://joss.theoj.org/papers/10.21105/joss.00640): Monte Carlo, summary statistics; input: data.frame, ML data; output: summary statistics
142. [`comorbidity`](https://joss.theoj.org/papers/10.21105/joss.00648): categorical variables, index; input: data.frame; output: model estimates
148. [`vtreat`](https://joss.theoj.org/papers/10.21105/joss.00584): ML, data preparation, missing value processing, categorical variables; input: data.frame; output: transformed data.frame
149. [`coalitions`](https://joss.theoj.org/papers/10.21105/joss.00606): uncertainty, redistribution, threshold, aggregation, Monte Carlo, Bayesian; input: data.frame; output: model prediction
151. [`ivprobit-1.0`](https://joss.theoj.org/papers/10.21105/joss.00523): probit model, misspecification; input: data.frame; output: model estimates, summary statistics
152. [`registr`](https://joss.theoj.org/papers/10.21105/joss.00557): functional data analysis, regression, warping; input: time series; output: model estimates
153. [`grapherator`](https://joss.theoj.org/papers/10.21105/joss.00528): networks, optimization, benchmarking; input: parameters; output: custom graph format
155. [`EFAshiny`](https://joss.theoj.org/papers/10.21105/joss.00567): EDA, factor analysis, visualization; input: data.frame; output: web-based visualization
157. [`reper`](https://joss.theoj.org/papers/10.21105/joss.00527): annotation, classification; input: application-specific data format; output: application-specific data format
158. [`psycho.R`](https://joss.theoj.org/papers/10.21105/joss.00470): reporting, model selection; input: data.frame, arbitrary data; output: summary statistics, visualization
160. [`origami`](https://joss.theoj.org/papers/10.21105/joss.00512): cross-validation, ML; input: data.frame; output: model prediction
162. [`GammaGompertzCR`](https://joss.theoj.org/papers/10.21105/joss.00216): survival, Bayesian, Monte Carlo; input: matrix; output: model prediction
166. [`BayesianNetwork`](https://joss.theoj.org/papers/10.21105/joss.00425): Bayesian, networks, ML; input: application-specific data format; output: summary statistics, visualization
174. [`hei`](https://joss.theoj.org/papers/10.21105/joss.00417): index; input: data.frame, application-specific data format; output: index scores
176. [`mcMST`](https://joss.theoj.org/papers/10.21105/joss.00374): networks, optimization; input: application-specific data format, graph; output: application-specific data format, graph
178. [`varistran`](https://joss.theoj.org/papers/10.21105/joss.00257): noise, variance; input: application-specific data format; output: summary statistics, model estimates, visualization
183. [`biotmle`](https://joss.theoj.org/papers/10.21105/joss.00295): optimization, covariates; input: application-specific data format; output: summary statistics, visualization
187. [`learningCurve`](https://joss.theoj.org/papers/10.21105/joss.00202): learning curve, estimates, aggregation; input: data.frame; output: model estimates
189. [`walkr`](https://joss.theoj.org/papers/10.21105/joss.00061): Monte Carlo, spatial, sampling, dimensionality reduction; input: matrix, data.frame; output: model estimates, summary statistics
192. [`remBoot`](https://joss.theoj.org/papers/10.21105/joss.00176): synthetic data, variance; input: data.frame; output: model estimates
194. [`KraljicMatrix`](https://joss.theoj.org/papers/10.21105/joss.00170): sensitivity, matrix algebra; input: data.frame; output: summary statistics, model prediction
195. [`nse`](https://joss.theoj.org/papers/10.21105/joss.00172): standard error, estimates; input: data.frame, numeric vectors; output: summary statistics
196. [`RiskPortfolios`](https://joss.theoj.org/papers/10.21105/joss.00171): risk, variance, covariance; input: data.frame, numeric vectors; output: model estimates, summary statistics
200. [`vbvs.concurrent`](https://joss.theoj.org/papers/10.21105/joss.00141): functional data analysis, Bayesian; input: data.frame, time series; output: model predictions, summary statistics
203. [`rucrdtw`](https://joss.theoj.org/papers/10.21105/joss.00100): warping, time series; input: time series; output: model estimates
204. [`CTLmapping`](https://joss.theoj.org/papers/10.21105/joss.00087): correlation, ANCOVA; input: application-specific data format; output: summary statistics
205. [`edarf`](https://joss.theoj.org/papers/10.21105/joss.00092): EDA, random forests, covariates, trees; input: data.frame; output: model prediction
213. [`gwdegree`](https://joss.theoj.org/papers/10.21105/joss.00036): networks, maximum entropy; input: application-specific data format; output: summary statistics

Then internally analyse this script to extract the statistical terms, and the input and output data types.

```{r}
x <- readLines ("categories.Rmd")
x <- x [grep ("[0-9]*\\. \\[", x)]
names <- vapply (x, function (i) {
                     res <- strsplit (i, "\\[\\`") [[1]]
                     strsplit (res, "\\`\\]") [[2]] [1] },
                     character (1),
                     USE.NAMES = FALSE)
terms <- lapply (x, function (i) {
                 res <- strsplit (i, ":\\s") [[1]] [2]
                 res <- strsplit (res, ";\\s") [[1]] [1] # rm "; input"
                 strsplit (res, ",\\s") [[1]]   })
input <- lapply (x, function (i) {
                     res <- strsplit (i, "input: ") [[1]] [2]
                     res <- strsplit (res, ";\\s") [[1]] [1]
                     if (grepl (",\\s", res))
                         res <- strsplit (res, ",\\s") [[1]]
                     return (res)   })
output <- lapply (x, function (i) {
                     res <- strsplit (i, "output: ") [[1]] [2]
                     if (grepl (",\\s", res))
                         res <- strsplit (res, ",\\s") [[1]]
                     return (res)   })
names (terms) <- names (input) <- names (output) <- names
knitr::kable (table (unlist (terms)))
```

Then convert to `visNetwork` nodes and edges tables:

```{r}
library (dplyr)
library (igraph)
nodes <- table (unlist (terms))
nodes <- data.frame (id = names (nodes),
                     label = names (nodes),
                     value = as.integer (nodes),
                     stringsAsFactors = FALSE)
edges <- lapply (terms, function (i) {
                     if (length (i) > 1) {
                         res <- sort (i)
                         n <- combn (seq_along (res), 2)
                         cbind (res [n [1, ]],
                                res [n [2, ]])
                     }  })
edges <- do.call (rbind, edges)
edges <- data.frame (from = edges [, 1],
                     to = edges [, 2],
                     stringsAsFactors = FALSE) %>%
    group_by (from, to) %>%
    summarise (width = length (from))
# Remove isolated edges:
# annotation, areal weights, binomial distribution, cubature, gene loci,
# generalized least squares, misspecification, classification, interpolation,
# over-dispersion, integration, random effects, probit model
cl <- graph_from_data_frame (edges) %>%
    clusters ()
out <- names (cl$membership [which (cl$membership != which.max (cl$csize))])
nodes <- nodes [which (!nodes$id %in% out), ]
edges <- edges [which (!(edges$from %in% out | edges$to %in% out)), ]
```

```{r visNetwork, eval = TRUE}
library (visNetwork)
visNetwork (nodes, edges)
```

or a manually-coded version, starting by converting node IDs and edge (from,
to) to integer values:
```{r node-edge-conversion}
nodes$id <- seq_along (nodes$id)
edges$from <- nodes$id [match (edges$from, nodes$label)]
edges$to <- nodes$id [match (edges$to, nodes$label)]
```


```{r vis-js-index-html}
index <- c ("<!--",
            "https://visjs.github.io/vis-network/examples/",
            "-->",
            "",
            "<!doctype html>",
            "<html>",
            "<head>",
            "   <meta charset=\"utf-8\" />",
            "   <title>Statistical Software Terms</title>",
            "   <script type=\"text/javascript\"",
            "       src=\"https://unpkg.com/vis-network/standalone/umd/vis-network.min.js\"></script>",
            "   <style type=\"text/css\">",
            "       html,",
            "       body {",
            "           font: 12pt arial;",
            "       }",
            "       #mynetwork {",
            "           width: \"100%\";",
            "           height: 90vh;",
            "           border: 1px solid lightgray;",
            "       }",
            "       code {",
            "           font-size: 15px;",
            "       }",
            "       indented {",
            "           margin-left: 30px;",
            "       }",
            "   </style>",
            "",
            "</head>",
            "",
            "<body>",
            "",
            "   <p>",
            "       Network connections between statistical terms used in JOSS papers.",
            "       Network magic courtesy of ",
            "       <a target=\"_blank\" rel=\"noopener noreferrer\"",
            "           href=\"https://github.com/visjs/vis-network\">",
            "           visjs/vis-network</a>, with",
            "       <a target=\"_blank\" rel=\"noopener noreferrer\"",
            "           href=\"https://github.com/ropenscilabs/statistical-software/blob/master/abstracts/categories.Rmd\">",
            "           code here</a>.",
            "   </p>",
            "   <div id=\"mynetwork\"></div>",
            "",
            "   <script src=\"mynetwork.js\" type=\"text/javascript\"></script>",
            "</body>",
            "</html>")
con <- file ("network-terms/index.html", "w")
writeLines (index, con = con)
close (con)
```

```{r vis-js-network}
out <- c ("var", 
          "   GREEN = \"rgba(0,255,150,0.8)\",",
          "   RED = \"rgba(255,50,0,0.8)\",",
          "   BLUE = \"rgba(0,150,255,0.8)\",",
          "   ORANGE = \"orange\";", "", "",
          "var nodes = new vis.DataSet([")
for (i in seq (nrow (nodes))) {
    out <- c (out,
              paste0 ("    { id: ", nodes$id [i], ","),
              paste0 ("      value: ", nodes$value [i], ","),
              paste0 ("      label: \"", nodes$label [i], "\" },"))
}
# rm terminal comma:
out [length (out)] <- gsub ("\\}\\,", "}", out [length (out)])
out <- c (out, "]);",
          "", "", "",
          "var edges = new vis.DataSet([")
for (i in seq (nrow (edges))) {
    out <- c (out,
              paste0 ("    { from: ", edges$from [i], ","),
              paste0 ("      to: ", edges$to [i], ","),
              paste0 ("      value: ", edges$width [i], " },"))
}
out [length (out)] <- gsub ("\\}\\,", "}", out [length (out)])

# remaining code:
out <- c (out,
          "]);", "", "", "",
          "var container = document.getElementById(\"mynetwork\");",
          "var data = {",
          "  nodes: nodes,",
          "  edges: edges",
          "};",
          "var options = {",
          "  nodes: {",
          "      shape: \"dot\",",
          "      scaling: {",
          "          customScalingFunction: function(min, max, total, value) {",
          "              return value / total;",
          "          },",
          "          min: 5,",
          "          max: 1000,",
          "      },",
          "      shadow: true",
          "  },",
          "  edges: {",
          "    shadow: true,",
          "    smooth: true",
          "  },",
          "  physics: {",
          "    barnesHut: { gravitationalConstant: -20000 },",
          "    stabilization: { iterations: 2500 }",
          "  }",
          "};",
          "",
          "var network = new vis.Network(container, data, options);",
          "",
          "network.on(\"click\", function(params) {",
          "    var node = nodes.get(params.nodes[0]);",
          "    if (typeof node.URL != 'undefined') {",
          "        window.open(node.URL,'_blank');",
          "    }",
          "});")
con <- file ("network-terms/mynetwork.js", "w")
writeLines (out, con = con)
close (con)
```

## Input-Output Relationships

Then do a similar network analysis relating input and output data forms

```{r io-nodes-edges}
library (dplyr)
terms <- table (c (unlist (input), unlist (output)))
nodes <- data.frame (id = seq_along (terms),
                     label = names (terms),
                     value = as.integer (terms),
                     stringsAsFactors = FALSE)
edges <- list ()
for (i in seq_along (input))
    edges [[i]] <- expand.grid (input [[i]], output [[i]])
edges <- do.call (rbind, edges) %>%
    data.frame (from = .$Var1,
                to = .$Var2) %>%
    group_by (from, to) %>%
    summarise (width = length (from))
edges$from <- nodes$id [match (edges$from, nodes$label)]
edges$to <- nodes$id [match (edges$to, nodes$label)]
# remove isolated edges:
cl <- graph_from_data_frame (edges) %>%
    clusters ()
out <- names (cl$membership [which (cl$membership != which.max (cl$csize))])
nodes <- nodes [which (!nodes$id %in% out), ]
edges <- edges [which (!(edges$from %in% out | edges$to %in% out)), ]
```


```{r vis-js-index-io-html}
index <- c ("<!--",
            "https://visjs.github.io/vis-network/examples/",
            "-->",
            "",
            "<!doctype html>",
            "<html>",
            "<head>",
            "   <meta charset=\"utf-8\" />",
            "   <title>Statistical Software I/O</title>",
            "   <script type=\"text/javascript\"",
            "       src=\"https://unpkg.com/vis-network/standalone/umd/vis-network.min.js\"></script>",
            "   <style type=\"text/css\">",
            "       html,",
            "       body {",
            "           font: 12pt arial;",
            "       }",
            "       #mynetwork {",
            "           width: \"100%\";",
            "           height: 90vh;",
            "           border: 1px solid lightgray;",
            "       }",
            "       code {",
            "           font-size: 15px;",
            "       }",
            "       indented {",
            "           margin-left: 30px;",
            "       }",
            "   </style>",
            "",
            "</head>",
            "",
            "<body>",
            "",
            "   <p>",
            "       Network connections between input and output data used in JOSS software",
            "       Network magic courtesy of ",
            "       <a target=\"_blank\" rel=\"noopener noreferrer\"",
            "           href=\"https://github.com/visjs/vis-network\">",
            "           visjs/vis-network</a>, with",
            "       <a target=\"_blank\" rel=\"noopener noreferrer\"",
            "           href=\"https://github.com/ropenscilabs/statistical-software/blob/master/abstracts/categories.Rmd\">",
            "           code here</a>.",
            "   </p>",
            "   <div id=\"mynetwork\"></div>",
            "",
            "   <script src=\"mynetwork.js\" type=\"text/javascript\"></script>",
            "</body>",
            "</html>")
con <- file ("network-io/index.html", "w")
writeLines (index, con = con)
close (con)
```

```{r vis-js-network-io}
out <- c ("var", 
          "   GREEN = \"rgba(0,255,150,0.8)\",",
          "   RED = \"rgba(255,50,0,0.8)\",",
          "   BLUE = \"rgba(0,150,255,0.8)\",",
          "   ORANGE = \"orange\";", "", "",
          "var nodes = new vis.DataSet([")
for (i in seq (nrow (nodes))) {
    out <- c (out,
              paste0 ("    { id: ", nodes$id [i], ","),
              paste0 ("      value: ", nodes$value [i], ","),
              paste0 ("      label: \"", nodes$label [i], "\" },"))
}
# rm terminal comma:
out [length (out)] <- gsub ("\\}\\,", "}", out [length (out)])
out <- c (out, "]);",
          "", "", "",
          "var edges = new vis.DataSet([")
for (i in seq (nrow (edges))) {
    out <- c (out,
              paste0 ("    { from: ", edges$from [i], ","),
              paste0 ("      to: ", edges$to [i], ","),
              paste0 ("      arrows: \"to\","),
              paste0 ("      value: ", edges$width [i], " },"))
}
out [length (out)] <- gsub ("\\}\\,", "}", out [length (out)])

# remaining code:
out <- c (out,
          "]);", "", "", "",
          "var container = document.getElementById(\"mynetwork\");",
          "var data = {",
          "  nodes: nodes,",
          "  edges: edges",
          "};",
          "var options = {",
          "  nodes: {",
          "      shape: \"dot\",",
          "      scaling: {",
          "          customScalingFunction: function(min, max, total, value) {",
          "              return value / total;",
          "          },",
          "          min: 5,",
          "          max: 500,",
          "      },",
          "      shadow: true",
          "  },",
          "  edges: {",
          "    shadow: true,",
          "    smooth: true",
          "  },",
          "  physics: {",
          "    barnesHut: { gravitationalConstant: -20000 },",
          "    stabilization: { iterations: 2500 }",
          "  }",
          "};",
          "",
          "var network = new vis.Network(container, data, options);",
          "",
          "network.on(\"click\", function(params) {",
          "    var node = nodes.get(params.nodes[0]);",
          "    if (typeof node.URL != 'undefined') {",
          "        window.open(node.URL,'_blank');",
          "    }",
          "});")
con <- file ("network-io/mynetwork.js", "w")
writeLines (out, con = con)
close (con)
```


