<!-- Edit the .Rmd not the .md file -->


## General Standards for Statistical Software {#general-standards}

These general standards, and all category-specific standards that follow, are
intended to serve as *recommendations* for best practices. Note in particular
that many standards are written using the word "*should*" in explicit
acknowledgement that adhering to such standards may not always be possible. All
standards phrased in these terms are intended to be interpreted as applicable
under such conditions as "*Where possible*", or "*Where applicable*".
Developers are requested to note any standards which they deem not applicable
to their software via the [`srr`
package](https://ropensci-review-tools.github.io/srr/), as described in
[Chapter 3](#pkgdev).

<details>
<summary>
These standards refer to <b>Data Types</b> as the fundamental types defined by the
<a href="https://cran.r-project.org/doc/manuals/R-lang.html">R language</a>
itself. Information on these types can be seen by clicking here.
</summary>
<p>

The [R language](https://cran.r-project.org/doc/manuals/R-lang.html) defines
the following data types:

- Logical
- Integer
- Continuous (`class = "numeric"` / `typeof = "double"`)
- Complex
- String / character

The base R system also includes what are considered here to be direct
extensions of fundamental types to include:

- Factor
- Ordered Factor
- Date/Time

The continuous type has a `typeof` of "double" because that represents the
storage mode in the C representation of such objects, while the `class` as
defined within R is referred to as "numeric". While `typeof` is not the same as
`class`, with reference to continuous variables, "numeric" may be considered
identical to "double" throughout.

The term "character" is interpreted here to refer to a vector each element of
which is an individual "character" object. The term "string" does not relate to
any official R nomenclature, but is used here to refer for convenience to
a character vector of length one; in other words, a "string" is the sole
element of a single-length "character" vector.

----

</p>
</details><br>

### Documentation

- [**G1.0**]{#G1_0} *Statistical Software should list at least one primary
  reference from published academic literature.*

We consider that statistical software submitted under our system will either
(i) implement or extend prior methods, in which case the *primary reference*
will be to the most relevant published version(s) of prior methods; or (ii) be
an implementation of some new method. In the second case, it will be expected
that the software will eventually form the basis of an academic publication.
Until that time, the most suitable reference for equivalent algorithms or
implementations should be provided.

- [**G1.1**]{#G1_1} *Statistical Software should document whether the
  algorithm(s) it implements are:*
    - *The first implementation of a novel algorithm*; or
    - *The first implementation within **R** of an algorithm which has
      previously been implemented in other languages or contexts*; or
    - *An improvement on other implementations of similar algorithms in **R***.

The second and third options additionally require references to comparable
algorithms or implementations to be documented somewhere within the software,
including references to all known implementations in other computer languages.
(A common location for such is a statement of "*Prior Art*" or similar at the
end of the main `README` document.)

- [**G1.2**]{#G1_2} *Statistical Software should include a* Life Cycle
  Statement *describing current and anticipated future states of development.*

We encourage these to placed within a repository's [`CONTRIBUTING.md`
file](https://devguide.ropensci.org/collaboration.html#contributing-guide), as
in [this
example](https://github.com/ecohealthalliance/fasterize/blob/master/CONTRIBUTING.md).
A simple *Life Cycle Statement* may be formed by selecting one of the following
four statements.

```
This package is

    - In a stable state of development, with minimal subsequent development
      envisioned.
    - In a stable state of development, with active subsequent development
      primarily in response to user feedback.
    - In a stable state of development, with some degree of active subsequent
      development as envisioned by the primary authors.
    - In an initially stable state of development, with a great deal of active
      subsequent development envisioned.
```


#### Statistical Terminology

- [**G1.3**]{#G1_3} *All statistical terminology should be clarified and
  unambiguously defined.*

Developers should not presume anywhere in the documentation of software that
specific statistical terminology may be “generally understood”, and therefore
not need explicit clarification. Even terms which many may consider
sufficiently generic as to not require such clarification, such as “null
hypotheses” or “confidence intervals”, will generally need explicit
clarification. For example, both the estimation and interpretation of
confidence intervals are dependent on distributional properties and associated
assumptions. Any particular implementation of procedures to estimate or report
on confidence intervals will accordingly reflect assumptions on distributional
properties (among other aspects), both the nature and implications of which
must be explicitly clarified.


#### Function-level Documentation

- [**G1.4**]{#G1_4} *Software should use
  [`roxygen2`](https://roxygen2.r-lib.org/) to document all functions.*
    - [**G1.4a**]{#G1_4a} *All internal (non-exported) functions should also be
      documented in standard [`roxygen2`](https://roxygen2.r-lib.org/) format,
      along with a final `@noRd` tag to suppress automatic generation of `.Rd`
      files.*

#### Supplementary Documentation

The following standards describe several forms of what might be considered
"Supplementary Material". While there are many places within an R package where
such material may be included, common locations include vignettes, or in
additional directories (such as `data-raw`) listed in `.Rbuildignore` to
prevent inclusion within installed packages.

Where software supports a publication, all claims made in the publication with
regard to software performance (for example, claims of algorithmic scaling or
efficiency; or claims of accuracy), the following standard applies:

- [**G1.5**]{#G1_5} *Software should include all code necessary to reproduce
  results which form the basis of performance claims made in associated
  publications.*

Where claims regarding aspects of software performance are made with respect to
other extant R packages, the following standard applies:

- [**G1.6**]{#G1_6} *Software should include code necessary to compare
  performance claims with alternative implementations in other R packages.*



### Input Structures

This section considers general standards for *Input Structures*. These
standards may often effectively be addressed through implementing class
structures, although this is not a general requirement. Developers are
nevertheless encouraged to examine the guide to [S3
vectors](https://vctrs.r-lib.org/articles/s3-vector.html#casting-and-coercion)
in the [`vctrs` package](https://vctrs.r-lib.org) as an example of the kind of
assurances and validation checks that are possible with regard to input data.
Systems like those demonstrated in that vignette provide a very effective way
to ensure that software remains robust to diverse and unexpected classes and
types of input data. Packages such
[`checkmate`](https://mllg.github.io/checkmate/index.html) enable direct and
simple ways to check and assert input structures. 

#### Uni-variate (Vector) Input

It is important to note for univariate data that single values in R are vectors
with a length of one, and that `1` is of exactly the same *data type* as `1:n`.
Given this, inputs expected to be univariate should:

- [**G2.0**]{#G2_0} *Implement assertions on lengths of inputs, particularly
  through asserting that inputs expected to be single- or multi-valued are
  indeed so.*
    - [**G2.0a**]{#G2_0a} Provide explicit secondary documentation of any
      expectations on lengths of inputs
- [**G2.1**]{#G2_1} *Implement assertions on types of inputs (see the initial
  point on nomenclature above).*
    - [**G2.1a**]{#G2_1a} *Provide explicit secondary documentation of
      expectations on data types of all vector inputs.*
- [**G2.2**]{#G2_2} *Appropriately prohibit or restrict submission of
  multivariate input to parameters expected to be univariate.*
- [**G2.3**]{#G2_3} *For univariate character input:*
    - [**G2.3a**]{#G2_3a} *Use `match.arg()` or equivalent where applicable to
      only permit expected values.*
    - [**G2.3b**]{#G2_3b} *Either: use `tolower()` or equivalent to ensure
      input of character parameters is not case dependent; or explicitly
      document that parameters are strictly case-sensitive.*
- [**G2.4**]{#G2_4} *Provide appropriate mechanisms to convert between
  different data types, potentially including:*
    - [**G2.4a**]{#G2_4a} *explicit conversion to `integer` via `as.integer()`*
    - [**G2.4b**]{#G2_4b} *explicit conversion to continuous via
      `as.numeric()`*
    - [**G2.4c**]{#G2_4c} *explicit conversion to character via
      `as.character()` (and not `paste` or `paste0`)*
    - [**G2.4d**]{#G2_4d} *explicit conversion to factor via `as.factor()`*
    - [**G2.4e**]{#G2_4e} *explicit conversion from factor via `as...()`
      functions*
- [**G2.5**]{#G2_5} *Where inputs are expected to be of `factor` type,
  secondary documentation should explicitly state whether these should be
  `ordered` or not, and those inputs should provide appropriate error or other
  routines to ensure inputs follow these expectations.*

A few packages implement R versions of "static type" forms common in other
languages, whereby the type of a variable must be explicitly specified prior to
assignment. Use of such approaches is encouraged, including but not restricted
to approaches documented in packages such as
[`vctrs`](https://vctrs.r-lib.org), or the experimental package
[`typed`](https://github.com/moodymudskipper/typed). One additional standard
for vector input is:

- [**G2.6**]{#G2_6} *Software which accepts one-dimensional input should ensure
  values are appropriately pre-processed regardless of class structures.*

The [`units` package](https://github.com/r-quantities/units/) provides a good
example, in creating objects that may be treated as vectors, yet which have
a class structure that does not inherit from the `vector` class. Using these
objects as input often causes software to fail. The `storage.mode` of the
underlying objects may nevertheless be examined, and the objects transformed or
processed accordingly to ensure such inputs do not lead to errors.


#### Tabular Input

This sub-section concerns input in "tabular data" forms, meaning the base
R forms `array`, `matrix`, and `data.frame`, and other forms and classes
derived from these. Tabular data generally have two dimensions, although may
have more (such as for `array` objects). There is a primary distinction within
R itself between `array` or `matrix` representations, and `data.frame` and
associated representations. The former are restricted to storing data of
a single uniform type (for example, all `integer` or all `character` values),
whereas `data.frame` as associated representations (generally) store each
column as a list item, allowing different columns to hold values of different
types. Further noting that a `matrix` may, [as of R version
4.0](https://developer.r-project.org/Blog/public/2019/11/09/when-you-think-class.-think-again/index.html),
be considered as a strictly two-dimensional array, tabular inputs for the
purposes of these standards are considered to imply data represented in one or
more of the following forms:

- `matrix` form when referring to specifically two-dimensional data of one
  uniform type
- `array` form as a more general expression, or when referring to data that are
  not necessarily or strictly two-dimensional
- `data.frame`
- Extensions such as
    - [`tibble`](https://tibble.tidyverse.org)
    - [`data.table`](https://rdatatable.gitlab.io/data.table)
    - domain-specific classes such as
      [`tsibble`](https://tsibble.tidyverts.org) for time series, or
      [`sf`](https://r-spatial.github.io/sf/) for spatial data.

Both `matrix` and `array` forms are actually stored as vectors with a single
`storage.mode`, and so all of the preceding standards **G2.0**--**G2.5** apply.
The other rectangular forms are not stored as vectors, and do not necessarily
have a single `storage.mode` for all columns. These forms are referred to
throughout these standards as "`data.frame`-type tabular forms", which may be
assumed to refer to data represented in either the `base::data.frame` format,
and/or any of the classes listed in the final of the above points.

General Standards applicable to software which is intended to accept any one or
more of these `data.frame`-type tabular inputs are then that:

- [**G2.7**]{#G2_7} *Software should accept as input as many of the above
  standard tabular forms as possible, including extension to domain-specific
  forms.*

Software need not necessarily test abilities to accept different types of
inputs, because that may require adding packages to the `Suggests` field of
a package for that purpose alone. Nevertheless, software which somehow uses
(through `Depends` or `Suggests`) any packages for representing tabular data
should confirm in tests the ability to accept these types of input.

- [**G2.8**]{#G2_8} *Software should provide appropriate conversion or dispatch
  routines as part of initial pre-processing to ensure that all other
  sub-functions of a package receive inputs of a single defined class or type.*
- [**G2.9**]{#G2_9} *Software should issue diagnostic messages for type
  conversion in which information is lost (such as conversion of variables from
  factor to character; standardisation of variable names; or removal of
  meta-data such as those associated with
  [`sf`-format](https://r-spatial.github.io/sf/) data) or added (such as
  insertion of variable or column names where none were provided).*

Note, for example, that an `array` may have column names which start with
numeric values, but that a `data.frame` may not.

```{r colnames}
x <- array (1, dim = c(1, 1), dimnames = list("1", "2")) # okay
print (x)
data.frame (x)
```

If `array` or `matrix` class objects are accepted as input, then **G2.8**
implies that routines should be implemented to check for such conversion of
column names.

The next standard concerns the following inconsistencies between three common
tabular classes in regard the column extraction operator, `[`.

```{r drop}
#| eval: false
x <- iris # data.frame from the datasets package
class (x)
#> [1] "data.frame"
class (x [, 1])
#> [1] "numeric"
class (x [, 1, drop = TRUE]) # default
#> [1] "numeric"
class (x [, 1, drop = FALSE])
#> [1] "data.frame"

x <- tibble::tibble (x)
class (x [, 1])
#> [1] "tbl_df"     "tbl"        "data.frame"
class (x [, 1, drop = TRUE])
#> [1] "numeric"
class (x [, 1, drop = FALSE]) # default
#> [1] "tbl_df"     "tbl"        "data.frame"

x <- data.table::data.table (x)
class (x [, 1])
#> [1] "data.table" "data.frame"
class (x [, 1, drop = TRUE]) # no effect
#> [1] "data.table" "data.frame"
class (x [, 1, drop = FALSE]) # default
#> [1] "data.table" "data.frame"
```

- Extracting a single column from a `data.frame` returns a `vector` by default,
  and a `data.frame` if `drop = FALSE`.
- Extracting a single column from a `tibble` returns a single-column `tibble`
  by default, and a `vector` if `drop = TRUE`. 
- Extracting a single column from a `data.table` always returns a `data.table`,
  and the `drop` argument has no effect.

Given such inconsistencies,

- [**G2.10**]{#G2_10} *Software should ensure that extraction or filtering of
  single columns from tabular inputs should not presume any particular default
  behaviour, and should ensure all column-extraction operations behave
  consistently regardless of the class of tabular data used as input.*

Adherence to the above standard **G2.8** will ensure that any implicitly or
explicitly assumed default behaviour will yield consistent results regardless
of input classes.

**Columns of tabular inputs**

The follow standards apply to `data.frame`-like tabular objects (including all
derived and otherwise compatible classes), and so do not apply to `matrix` or
`array` objects.

- [**G2.11**]{#G2_11} *Software should ensure that `data.frame`-like tabular
  objects which have columns which do not themselves have standard class
  attributes (typically, `vector`) are appropriately processed, and do not
  error without reason. This behaviour should be tested. Again, columns created
  by the [`units` package](https://github.com/r-quantities/units/) provide
  a good test case.*
- [**G2.12**]{#G2_12} *Software should ensure that `data.frame`-like tabular
  objects which have list columns should ensure that those columns are
  appropriately pre-processed either through being removed, converted to
  equivalent vector columns where appropriate, or some other appropriate
  treatment such as an informative error. This behaviour should be tested.*

#### Missing or Undefined Values

- [**G2.13**]{#G2_13} *Statistical Software should implement appropriate checks
  for missing data as part of initial pre-processing prior to passing data to
  analytic algorithms.*
- [**G2.14**]{#G2_14} *Where possible, all functions should provide options for
  users to specify how to handle missing (`NA`) data, with options minimally
  including:*
  - [**G2.14a**]{#G2_14a} *error on missing data*
  - [**G2.14b**]{#G2_14b} *ignore missing data with default warnings or
    messages issued*
  - [**G2.14c**]{#G2_14c} *replace missing data with appropriately imputed
    values*
- [**G2.15**]{#G2_15} *Functions should never assume non-missingness, and
  should never pass data with potential missing values to any base routines
  with default `na.rm = FALSE`-type parameters (such as
  [`mean()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/mean.html),
  [`sd()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/sd.html) or
  [`cor()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cor.html)).*
- [**G2.16**]{#G2_16} *All functions should also provide options to handle
  undefined values (e.g., `NaN`, `Inf` and `-Inf`), including potentially
  ignoring or removing such values.*

### Algorithms

- [**G3.0**]{#G3_0} *Statistical software should never compare floating point
  numbers for equality. All numeric equality comparisons should either ensure
  that they are made between integers, or use appropriate tolerances for
  approximate equality.*

This standard applies to all computer languages included in any package. In R,
values can be affirmed to be integers through `is.integer()`, or asserting that
the `storage.mode()` of an object is "integer". One way to compare numeric
values with tolerance is with the [`all.equal()`
function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/all.equal.html),
which accepts an additional `tolerance` parameter with a default for `numeric`
comparison of `sqrt(.Machine$double.eps)`, which is typically around e(-8--10).
In other languages, including C and C++, comparisons of floating point numbers
are commonly implemented by conditions such as `if (abs(a - b) < tol)`, where
`tol` specifies the tolerance for equality.

Importantly, R functions such as
[`duplicated()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/duplicated.html)
and
[`unique()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/unique.html)
rely on equality comparisons, and this standard extends to require that
software should not apply any functions which themselves rely on equality
comparisons to floating point numbers.

- [**G3.1**]{#G3_1} *Statistical software which relies on covariance
  calculations should enable users to choose between different algorithms for
  calculating covariances, and should not rely solely on covariances from the
  `stats::cov` function.*
    - [**G3.1a**]{#G3_1a} *The ability to use arbitrarily specified covariance
      methods should be documented (typically in examples or vignettes).*

Estimates of covariance can be very sensitive to outliers, and a variety of
methods have been developed for "robust" estimates of covariance, implemented
in such packages as [`rms`](https://cran.r-project.org/package=rms),
[`robust`](https://cran.r-project.org/package=robust), and
[`sandwich`](https://cran.r-project.org/package=sandwich). Adhering to this
standard merely requires an ability for a user to specify a particular
covariance function, such as through an additional parameter. The `stats::cov`
function can be used as a default, and additional packages such as the three
listed here need not necessarily be listed as `Imports` to a package.


### Output Structures

- [**G4.0**]{#G4_0} *Statistical Software which enables outputs to be written
  to local files should parse parameters specifying file names to ensure
  appropriate file suffices are automatically generated where not provided.*

### Testing 

All packages should follow rOpenSci standards on
[testing](https://devguide.ropensci.org/building.html#testing) and [continuous
integration](https://devguide.ropensci.org/ci.html), including aiming for high
test coverage. Extant R packages which may be useful for testing include
[`testthat`](https://testthat.r-lib.org),
[`tinytest`](https://github.com/markvanderloo/tinytest),
[`roxytest`](https://github.com/mikldk/roxytest), and
[`xpectr`](https://github.com/LudvigOlsen/xpectr).

#### Test Data Sets

- [**G5.0**]{#G5_0} *Where applicable or practicable, tests should use standard
  data sets with known properties (for example, the [NIST Standard Reference
  Datasets](https://www.itl.nist.gov/div898/strd/), or data sets provided by
  other widely-used R packages).*
- [**G5.1**]{#G5_1} *Data sets created within, and used to test, a package
  should be exported (or otherwise made generally available) so that users can
  confirm tests and run examples.*

#### Responses to Unexpected Input

- [**G5.2**]{#G5_2} *Appropriate error and warning behaviour of all functions
  should be explicitly demonstrated through tests. In particular,*
    - [**G5.2a**]{#G5_2a} *Every message produced within R code by `stop()`,
      `warning()`, `message()`, or equivalent should be unique*
    - [**G5.2b**]{#G5_2b} *Explicit tests should demonstrate conditions which
      trigger every one of those messages, and should compare the result with
      expected values.*
- [**G5.3**]{#G5_3} *For functions which are expected to return objects
  containing no missing (`NA`) or undefined (`NaN`, `Inf`) values, the absence
  of any such values in return objects should be explicitly tested.*

#### Algorithm Tests

For testing _statistical algorithms_, tests should include tests of the
following types:

- [**G5.4**]{#G5_4} **Correctness tests** *to test that statistical algorithms
  produce expected results to some fixed test data sets (potentially through
  comparisons using binding frameworks such as
  [RStata](https://github.com/lbraglia/RStata)).*
    - [**G5.4a**]{#G5_4a} *For new methods, it can be difficult to separate out
      correctness of the method from the correctness of the implementation, as
      there may not be reference for comparison.  In this case, testing may be
      implemented against simple, trivial cases or against multiple
      implementations such as an initial R implementation compared with results
      from a C/C++ implementation.*
    - [**G5.4b**]{#G5_4b} *For new implementations of existing methods,
      correctness tests should include tests against previous implementations.
      Such testing may explicitly call those implementations in testing,
      preferably from fixed-versions of other software, or use stored outputs
      from those where that is not possible.*
    - [**G5.4c**]{#G5_4c} *Where applicable, stored values may be drawn from
      published paper outputs when applicable and where code from original
      implementations is not available*
- [**G5.5**]{#G5_5} *Correctness tests should be run with a fixed random seed*
- [**G5.6**]{#G5_6} **Parameter recovery tests** *to test that the
  implementation produce expected results given data with known properties.
  For instance, a linear regression algorithm should return expected
  coefficient values for a simulated data set generated from a linear model.*
    - [**G5.6a**]{#G5_6a} *Parameter recovery tests should generally be
      expected to succeed within a defined tolerance rather than recovering
      exact values.*
    - [**G5.6b**]{#G5_6b} *Parameter recovery tests should be run with multiple
      random seeds when either data simulation or the algorithm contains
      a random component. (When long-running, such tests may be part of an
      extended, rather than regular, test suite; see G4.10-4.12, below).*

Note that authors should ensure that they use [at least v3 of the `testthat`
package](https://testthat.r-lib.org), which introduced a
[`testthat_tolerance()`](https://testthat.r-lib.org/reference/compare.html),
defaulting to the value defined by `base::all_equal()` of
`sqrt(.Machine$double.eps)` on all `expect_equal()` expectations.

- [**G5.7**]{#G5_7} **Algorithm performance tests** *to test that
  implementation performs as expected as properties of data change.  For
  instance, a test may show that parameters approach correct estimates within
  tolerance as data size increases, or that convergence times decrease for
  higher convergence thresholds.*
- [**G5.8**]{#G5_8} **Edge condition tests** *to test that these conditions
  produce expected behaviour such as clear warnings or errors when confronted
  with data with extreme properties including but not limited to:*
    - [**G5.8a**]{#G5_8a} *Zero-length data*
    - [**G5.8b**]{#G5_8b} *Data of unsupported types (e.g., character or
      complex numbers in for functions designed only for numeric data)*
    - [**G5.8c**]{#G5_8c} *Data with all-`NA` fields or columns or all
      identical fields or columns*
    - [**G5.8d**]{#G5_8d} *Data outside the scope of the algorithm (for
      example, data with more fields (columns) than observations (rows) for
      some regression algorithms)*
- [**G5.9**]{#G5_9} **Noise susceptibility tests** *Packages should test for
  expected stochastic behaviour, such as through the following conditions:*
    - [**G5.9a**]{#G5_9a} *Adding trivial noise (for example, at the scale of
      `.Machine$double.eps`) to data does not meaningfully change results*
    - [**G5.9b**]{#G5_9c} *Running under different random seeds or initial
      conditions does not meaningfully change results*

#### Extended tests

Thorough testing of statistical software may require tests on large data sets, 
tests with many permutations, or other conditions leading to long-running
tests. In such cases it may be neither possible nor advisable to execute tests
continuously, or with every code change. Software should nevertheless test any
and all conditions regardless of how long tests may take, and in doing so
should adhere to the following standards:

- [**G5.10**]{#G5_10} *Extended tests should included and run under a common
  framework with other tests but be switched on by flags such as as
  a `<MYPKG>_EXTENDED_TESTS="true"` environment variable.*
    - The extended tests can be then run automatically by GitHub Actions for 
      example by adding the following to the `env` section of the workflow:
      
      `MYPKG_EXTENDED_TESTS: ${{contains(github.event.head_commit.message, 'run-extended')}}`
      
      Extended tests will then be run in response to any commit message which contains the phrase `run-extended`.
- [**G5.11**]{#G5_11} *Where extended tests require large data sets or other
  assets, these should be provided for downloading and fetched as part of the
  testing workflow.*
    - [**G5.11a**]{#G5_11a} *When any downloads of additional data necessary
      for extended tests fail, the tests themselves should not fail, rather be
      skipped and implicitly succeed with an appropriate diagnostic message.*
- [**G5.12**]{#G5_12} *Any conditions necessary to run extended tests such as
  platform requirements, memory, expected runtime, and artefacts produced that
  may need manual inspection, should be described in developer documentation
  such as a `CONTRIBUTING.md` or `tests/README.md` file.*
