<!-- Edit the .Rmd not the .md file -->

<!-- TODO
- revisit regression standards and use a a guide for how to write/adapt actual
  algorithmic standards for spatial software.
- permutation-type statistics which need random seeds
- clustering
- spatial as a context for representating data which is then passed through to
  algorithms which themselves might be subject to additional standards.
- Kernel density estimates: What are default values for parameters? How are
  they justified? How can they be controlled?
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
                      echo = TRUE,
                      out.width = "100%",
                      collapse = TRUE,
                      comment = "#>"
)
```

## Spatial Software

Standards for spatial software begin with a consideration and standardisation
of domains of applicability. Following that we proceed to standards according
to which spatial software is presumed to perform one or more of the following
steps:

1. Accept and validate input data
2. Apply one or more analytic algorithms
3. Return the result of that algorithmic application
4. Offer additional functionality such as printing or summarising return results
5. Testing

Each standard for spatial software is prefixed with "**SP**".

### Spatial Domains

Many developers of spatial software in R, including those
responsible for the CRAN Task view on ["Analysis of Spatial
Data"](https://cran.r-project.org/web/views/Spatial.html), have been primarily
focussed on geographic data; that is, data quantifying positions, structures,
and relationships on the Earth and other planets. Spatial analyses are
nevertheless both broader and more general than geography alone. In particular,
spatial software may be *geometric* -- that is, concerned with 
positions, structures, and relationships in space in any general or specific
sense, not necessarily confined to geographic systems alone.

It is important to distinguish these two domains because many algorithms and
procedures devised in one of these two domains are not necessarily (directly)
applicable in the other, most commonly because geometric algorithms presume
space to be rectilinear, while geographic algorithms (generally) presume it be
have a specific curvilinear form. Algorithms designed for rectilinear space may
not be directly applicable in curvilinear space, and vice-versa.

Moreover, spatial software and algorithms might be intended to apply in spaces
of arbitrary dimensionality. The phrase "rectilinear" refers to any space of
arbitrary dimensionality in which all dimensions are described by straight
lines; dimensions in a curvilinear space or arbitrary dimensionality are
described by curved lines. A planar geometry is a two-dimensional rectilinear
space; a spherical geometry is a two- (or maybe three-)dimensional curvilinear
space.

One of the earliest and still most widely used R spatial packages,
[`spatstat`](https://cran.r-project.org/web/packages/spatstat/) (first released
2002), describes itself as, "[f]ocused mainly on two-dimensional point
patterns, including multitype/marked points, in any spatial region." Routines
from this package are thus generally applicable to two-dimensional data only, 
while the final phrase might be interpreted to indicate a comprehensive
generality. Nevertheless, almost all of the routines assume rectilinear (in
this case, planar) space and may not necessarily give accurate results when
applied in curvilinear space. 

These considerations motivate the first standard for spatial software:

- **SP1.0** *Spatial software should explicitly indicate its domain of
  applicability, and in particular distinguish whether the software may be
  applied in rectilinear/geometric domains, curvilinear/geographic domains, or
  both.*

We encourage the use of clear and unambiguous phrases such as "planar",
"spherical", "rectilinear" or "curvilinear", along with clear indications of
dimensionality such as "two-" or "three-dimensional." Concepts of
dimensionality should be interpreted to refer explicitly to the dimensionality
of independent spatial coordinates. Elevation is a third spatial dimension,
while other attributes measured at spatial locations do not represent
additional dimensions. 

- **SP1.1** *Spatial software should explicitly indicate its dimensional domain
  of applicability, in particular through identifying whether it is applicable
  to two or three dimensions only, or whether there are any other restrictions
  on dimensionality.*

These considerations of domains of applicability permeate much of the ensuring
standards, which distinguish "geometric software" from "geographic software",
where these phrases are to be interpreted as shorthand references to software
intended for use in the respective domains.


### Input data structures and validation

Input validation is an important software task, and an important part of our
standards. While there are many ways to approach validation, the class systems
of R offer a particularly convenient and effective means. For Spatial
Software in particular, a range of class systems have been developed, for which
we refer to the CRAN Task view on ["Analysis of Spatial
Data"](https://cran.r-project.org/web/views/Spatial.html). Software which uses
and relies on defined classes can often validate input through affirming
appropriate class(es). Software which does not use or rely on class systems
will generally need specific routines to validate input data structures.

These standards for Spatial Software do not explicitly recommend or require
that class systems be used, unlike our standards for [Time-Series
Software](https://ropenscilabs.github.io/statistical-software-review-book/standards.html#time-series-software)
which do recommend the use of explicit class systems. The latter recommendation
is because time series data can often be represented in a single dimension only
in which developers may assume that data are temporally regular. Representation
as a simple vector negates any ability to assert or test whether input data are
indeed regular or not, whereas class systems for time series data directly
enable such assertions. In contrast, spatial data are both inherently
multi-dimensional and rarely regular, and so properties of input data can be
more readily asserted and tested even without explicit class systems. (An
important form of regular spatial data is raster data, for which the following
standard also applies.)

- **SP2.0** *Spatial software should adhere to one or the other of the
  following standards, dependent on whether input data may be generic
  (non-class-based), or of one or more specific classes.*
    - **SP2.0a** *Spatial Software intended for use in geographical (or more
      general curvilinear domains) should only accept input data of one or more
      classes explicitly developed to represent such data.*
    - **SP2.0b** *Spatial Software which is intended for use in rectilinear
      domains and which accepts generic (non-class-based) inputs should assert
      that input data are not obviously curvilinear or spherical, commonly
      through asserting that coordinate columns are not labelled any variant of
      "longitude" or "latitude", and should implement an appropriate response
      when such assertions fail.*


A failure to assert that input columns do not describe curvilinear or spherical
data may have different implications for different software, with actions
ranging from automatic transformation to planar form to outright errors. The
following illustrate warning messages issued by the [`sf`
package](https://cran.r-project.org/package=sf) when operations which assume
planar data are applied to spherical data (defined via `crs = 4326`).
Both examples impose a spherical coordinate reference system (`crs`) on the
data before applying operations which assume planar data, with the developers
opting in both cases to issue warnings.

```{r sf-warnings, eval = FALSE}
library (sf)
p <- rbind (c (0, 0), c (1, 0), c (3, 2), c (2, 4), c (1, 4), c (0, 0))
pol <- st_sfc (st_polygon (list (p)), crs = 4326)
cent <- st_centroid (pol)
#> Warning in st_centroid.sfc(pol): st_centroid does not give correct centroids for
#> longitude/latitude data

pt <- st_sfc (st_point (c (0, 0)), crs = 4326)
buf <- st_buffer (pt, 1)
#> Warning in st_buffer.sfc(pt, 1): st_buffer does not correctly buffer longitude/
#> latitude data
```

These warnings are able to be appropriately issued because the functions
[`st_centroid()` and
`st_buffer()`](https://r-spatial.github.io/sf/reference/geos_unary.html) only
accept as input an "object of class `sfg`, `sfc`, or `sf`," with these classes
enabling assertion of cases in which coordinates reflect curvilinear (more
specifically in this case, [ellipsoidal](https://epsg.io/4326)) geometries.

**Spatial Workflows, Packages, and Classes**

Spatial software encompasses an enormous diversity, yet workflows implemented
by spatial software often share much in common. In particular, coordinate
reference systems used to precisely relate pairs of coordinates to precise
locations in a curvilinear space, and in particular to the Earth's ellipsoid,
need to be able to be compared and transformed regardless of the specificities
of individual software. This ubiquitous need has fostered the development of
the [`PROJ` library](https://proj.org/) for representing and transforming
spatial coordinates. Several other libraries have been built on top or or
alongside that, notably including the [`GDAL` ("Geospatial Data Abstraction
Library")](https://gdal.org) and [`GEOS` ("Geometry Engine, Open
Source")](https://trac.osgeo.org/geos/) libraries. These libraries are used by,
and integrated within, most geographical spatial software commonly used today,
and will likely continue to be used.

While not a standard in itself, it is expected that spatial software should
not, absent very convincing and explicit justification, attempt to reconstruct
aspects of these generic libraries. Given that, the following standards aim to
ensure that spatial software remains as compatible as possible with workflows
established by preceding packages which have aimed to expose and integrate as
much of the functionality of these generic libraries as possible. The use of
specific class systems for spatial data, and the workflows encapsulated in
associated packages, ensures maximal ongoing compatibility with these
libraries and with spatial workflows in general.

Notable class systems and associated packages in R include
[`sp`](https://cran.r-project.org/package=sp),
[`sf`](https://cran.r-project.org/package=sf), and
[`raster`](https://rspatial.org/raster/), and more recent extensions such as
[`stars`](https://cran.r-project.org/package=stars),
[`terra`](https://rspatial.org/terra), and
[`s2`](https://r-spatial.github.io/s2/). With regard to these packages, the
following single standard applies:

- **SP2.1** *Spatial Software which uses either the
  [`sf`](https://cran.r-project.org/package=sf) of
  [`sp`](https://cran.r-project.org/package=sp) 
  class systems for representing geographical data should either*
    - **SP2.1a** *Use [`sf`](https://cran.r-project.org/package=sf) rather
      than [`sp`](https://cran.r-project.org/package=sp),* or
    - **SP2.1b** *Explicitly justify why
      [`sp`](https://cran.r-project.org/package=sp) is used.*

An example of a valid explicit reason to use
[`sp`](https://cran.r-project.org/package=sp) would be to ensure compatibility
with some other dependent package which exclusively relies on input data in
that format.

More generally,

- **SP2.2** *Geographical Spatial Software should ensure maximal compatibility
  with established packages and workflows, minimally through:*
    - **SP2.2a** *Clear and extensive documentation demonstrating how routines
      from that software may be embedded within, or otherwise adapted to,
      workflows which rely on these established packages; and*
    - **SP2.2b** *Tests which clearly demonstrate that routines from that
      software may be successfully translated into forms and workflows which
      rely on these established packages.*

This standard is further refined in a number of subsequent standards concerning
documentation and testing.

**Coordinate Reference Systems**

As described above, one of the primary reasons for the development of classes
in Spatial Software is to represent the coordinate reference systems in which
data are represented, and to ensure compatibility with the [`PROJ`
system](https://proj.org/) and other generic spatial libraries. The
[`PROJ`](https://proj.org/) standards and associated software library have been
recently (2020) updated (to version number 7) with "breaking changes" that are
not backwards-compatible with previous versions, and in particular with the
long-standing version 4. The details and implications of these changes within
the context of spatial software in R can be examined in [this blog
entry](https://www.r-spatial.org//r/2020/03/17/wkt.html) on
[`r-spatial.org`](https://r-spatial.org), and in [this
vignette](https://cran.r-project.org/web/packages/rgdal/vignettes/PROJ6_GDAL3.html)
for the [`rgdal` package](https://cran.r-project.org/web/packages/rgdal/). The
"breaking" nature of these updates partly reflects analogous "breaking changes"
associated with updates in the ["Well-Known Text"
(WKT)](http://docs.opengeospatial.org/is/12-063r5/12-063r5.html) system for
representing coordinate reference systems. 

The following standard applies to software which directly or indirectly relies
on geographic data which uses or relies upon coordinate reference systems.

- **SP2.3** *Geographical Spatial Software should be compliant with version 6
  (and, ideally 7) of* [`PROJ`](https://proj.org/), *and with* `wkt2`
  *representations. The primary implications, described in detail in the articles
  linked to above, are that:*
    - **SP2.3a** *Software should not accept so-called "PROJ4-strings"
      previously used to specify coordinate reference systems.*
    - **SP2.3b** *Documentation should explicitly clarify whether, and under
      which conditions, geographical coordinates are expected to be
      longitude-latitude or latitude-longitude.*


**General Input Structures**

New spatial software may nevertheless eschew these prior packages and classes
in favour of implementing new classes. Whether or not prior classes are used or
expected, geographic software should accord as much as possible with the
principles of these prior systems by according with the following standards:

- **SP2.4** *Class systems for input data must contain meta data on associated
  coordinate reference systems.*
    - **SP2.4a** *Software should provide an ability to convert objects in any
      new class systems into representations of pre-existing classes such as
      those listed above.*
- **SP2.5** *Spatial Software should explicitly document the types and classes
  of input data able to be passed to each function.*
- **SP2.6** *Spatial Software should accept input data in as many specific
  spatial classes as possible.*
- **SP2.7** *Spatial Software should implement validation routines to confirm
  that inputs are of acceptable classes (or represented in otherwise
  appropriate ways for software which does not use class systems).*
- **SP2.8** *Spatial Software should implement a single pre-processing routine
  to validate input data, and to appropriately transform it to a single uniform
  type to be passed to all subsequent data-processing functions.*
- **SP2.9** *The pre-processing function described above should maintain all
  metadata attributes of input data.*

#### Spatial Scales and Units

Many spatial routines are intended to accept spatial data measured on some
specific scale, yet data on actual scales of measurement are commonly
discarded. The [`units` package](https://github.com/r-quantities/units/) has been developed to
attribute SI units to R vectors. While we do not suggest via explicit standards
that this package be used in Spatial Software, the following standard should
nevertheless be adhered to:

-   **SP2.10** *Spatial software should accept inputs defined via the* [`units`
    package](https://github.com/r-quantities/units/).


### Algorithms

The following standards will be conditionally applicable to some but not all
spatial software. Procedures for standards deemed not applicable to
a particular piece of software are described in the [R package of this
project](https://github.com/ropenscilabs/statistical-software-review).

- **SP3.0** *Spatial software which considers spatial neighbours should enable
  user control over neighbourhood forms and sizes. In particular:*
    - **SP3.0a** *Neighbours (able to be expressed) on regular grids should be
      able to be considered in both rectangular only, or rectangular and
      diagonal (respectively "rook" and "queen" by analogy to chess.*
    - **SP3.0b** *Neighbourhoods in irregular spaces should be minimally able
      to be controlled via an integer number of neighbours, an area (or
      equivalent distance defining an area) in which to include neighbours, or
      otherwise equivalent user-controlled value.*
- **SP3.1** *Spatial software which considers spatial neighbours should enable
  neighbour contributions to be weighted by distance (or other weighting
  variable), and not rely on a uniform-weight rectangular cut-off.*
- **SP3.2** *Spatial software which relies on sampling from input data (even if
  only of spatial coordinates) should enable sampling procedures to be based on
  local spatial densities of those input data.*

Algorithms for spatial software are often related to other categories of
statistical software, and it is anticipated that spatial software will commonly
also be subject to standards from these other categories. Nevertheless, because
spatial analyses frequently face unique challenges, some of these
category-specific standards also have extension standards when applied to
spatial software. The following standards will be applicable for any spatial
software which also fits any of the other listed categories of statistical
software.

**Regression Software**

- **SP3.3** *Spatial regression software should explicitly quantify and
  distinguish autocovariant or autoregressive processes from those covariant or
  regressive processes not directly related to spatial structure alone.*

**Unsupervised Learning Software**

The following standard applies to any spatial unsupervised learning software
which uses clustering algorithms.

- **SP3.4** *Spatial clustering should not use standard non-spatial clustering
  algorithms in which spatial proximity is merely represented by an additional
  weighting factor. Rather, clustering schemes should be derived from
  explicitly spatial algorithms.*

**Machine Learning Software**

One common application in which machine learning algorithms are applied to
spatial software is in analyses of raster images. The first of the following
standards applies because the individual cells or pixels of these raster images
represent fixed spatial coordinates. (This standard also renders **ML2.1**
inapplicable).


- **SP3.5** *Spatial machine learning software should ensure that broadcasting
  procedures for reconciling inputs of different dimensions are **not**
  applied*.
- **SP3.6** *Spatial machine learning software should ensure that test and
  training data are spatially distinct, and not simply sampled uniformly from
  a common region.*

The latter standard, **SP3.6**, is commonly met by applying some form of
spatial partitioning to data, and using spatially distinct partitions to define
test and training data.

      
### Return Results

For (functions within) Spatial Software which return spatial data:

- **SP4.0** *Return values should either:*
    - **SP4.0a** *Be in same class as input data, or*
    - **SP4.0b** *Be in a unique, preferably class-defined, format.*
- **SP4.1** *Any units included as attributes of input data should also be included
  within return values.*
- **SP4.2** *The type and class of all return values should be explicitly documented.*

For (functions within) Spatial Software which return data other than direct
series:

- **SP4.3** *Return values should explicitly include all appropriate units*


### Visualization

Spatial Software which returns objects in a custom class structure should:

- **SP5.0** *Implement default `plot` methods for any implemented class system.*
- **SP5.1** *Default to placing the "longitude" or "x" (or equivalent) variable
  on the horizontal axis.*
- **SP5.2** *Ensure that any spatial units associated with input coordinates,
  and maintained in the return object according to* **SP4.1**, *are  printed by
  default on the axes.*

Spatial Software which returns objects with geographical coordinates should:

- **SP5.3** *Offer an ability to generate interactive (generally `html`-based)
  visualisations of results.*

### Testing

The following standards apply to all Spatial Software which is intended or able
to be applied to data represented in curvilinear systems, notably including all
geographical data. The only Spatial Software to which the following standards
do not (necessarily) apply would be software explicitly intended to be applied
exclusively to rectilinear spatial data, and which ensured appropriate
rejection of curvilinear data according to **SP2.0b**.

**Round-Trip Tests**

- **SP6.0** *Software which implements routines for transforming coordinates of
  input data should include tests which demonstrate ability to recover the
  original coordinates.*
- **SP6.1** *All functions which can be applied to both rectilinear and
  curvilinear data should be tested through application to both.*
    - **SP6.1a** *Functions which may yield inaccurate results when applied to
      data in one or the other forms (such as the preceding examples of
      centroids and buffers from ellipsoidal data) should test that results
      from inappropriate application of those functions are indeed less
      accurate.*
    - **SP6.1b** *Functions which yield accurate results regardless of whether
      input data are rectilinear or curvilinear should demonstrate equivalent
      accuracy in both cases, and should also demonstrate how equivalent
      results may be obtained through first explicitly transforming input data.*

**Extreme Geographical Coordinates**

- **SP6.2** *Geographical Software should include tests with extreme
  geographical coordinates, minimally including extension to polar extremes of
  +/-90 degrees.*

While such tests should generally confirm that software generates reliable
results to such extreme coordinates, software which is unable to generate
reliable results to such inputs should nevertheless include tests to indicate
both approximate bounds of reliability, and the expected characteristics of
unreliable results.

The remaining standards for testing Spatial Software extend directly from the
preceding Algorithmic Standards (**SP3**), with the same sub-section headings
used here.

- **SP6.3** *Spatial Software which considers spatial neighbours should
  explicitly test all possible ways of defining them, and should explicitly
  compare quantitative effects of different ways of defining neighbours.*
- **SP6.4** *Spatial Software which considers spatial neighbours should
  explicitly test effects of different schemes to weight neighbours by spatial
  proximity.*

**Unsupervised Learning Software**

- **SP6.5** *Spatial Unsupervised Learning Software which uses clustering
  algorithms should implement tests which explicitly compare results with
  equivalent results obtained with a non-spatial clustering algorithm.*

**Machine Learning Software**

- **SP6.6** *Spatial Machine Learning Software should implement tests which
  explicitly demonstrate the detrimental consequences of sampling test and
  training data from the same spatial region, rather than from spatially
  distinct regions.
